---
title: "Relationship between Transmission Type and Fuel Efficiency"
author: "F. Alex Crofut"
date: "Saturday, April 30, 2016"
output: html_document
---

##Executive Summary
Is an automatic or manual transmission better for MPG
Quantify the MPG difference between automatic and manual transmissions


##Key Assumptions



##Exploratory Data Analysis
###Required Packages
```{r packages}
if (!"datasets" %in% installed.packages()) install.packages("datasets")
library("datasets")
if (!"ggplot2" %in% installed.packages()) install.packages("ggplot2")
library("ggplot2")
if (!"gridExtra" %in% installed.packages()) install.packages("gridExtra")
library("gridExtra")
if (!"caret" %in% installed.packages()) install.packages("caret")
library("caret")

```

##Data Structure and Processing
###Structure
```{r structure}
data(mtcars)
str(mtcars)
summary(mtcars)
```

###Processing
```{r processing}
# replace 0&1 with "automatic" and "manual" in new dataframe
mtcars2 <- mtcars
mtcars2$am[mtcars2$am == 0] <- "automatic"
mtcars2$am[mtcars2$am == 1] <- "manual"


```

###Data Exploration
```{r exploration}
p1 <- ggplot(aes(x = am, y = mpg), data = mtcars2) +
      geom_boxplot(fill = "blue") +
      labs(x = "Transmission Type", y = "Miles per Gallon") + 
      ggtitle("Miles per Gallon by Transmission Type")

p2 <- featurePlot(x = mtcars2[,c(1,3,6,11)], y = mtcars2$am, plot = "pairs", auto.key = list(columns = 2))

p3 <- ggplot(mtcars2, aes(x = mpg)) + geom_histogram(binwidth = .5, colour="black", fill = "blue") + 
        facet_wrap(~ am)

grid.arrange(p1, p3, ncol = 2)
p2

```

##Models

###Assumptions

* Population is independent and identically distributed.
* Measurement error is accounted for by significant digits.
* A 95% confidence interval is sufficient to determine significance.

###Strategy for model selection and Fitting Multiple Models


###Residual Plot and Diagnostics


###Quantify Uncertainty



##Conclusions