---
title: "Relationship between Transmission Type and Fuel Efficiency"
author: "F. Alex Crofut"
date: "Saturday, April 30, 2016"
output: html_document
---

##Executive Summary
Is an automatic or manual transmission better for MPG
Quantify the MPG difference between automatic and manual transmissions

##Assumptions
* Population is independent and identically distributed.
* Measurement error is accounted for by significant digits.
* A 95% confidence interval is sufficient to determine significance.

##Exploratory Data Analysis
###Required Packages
```{r packages}
if (!"datasets" %in% installed.packages()) install.packages("datasets")
library("datasets")
if (!"ggplot2" %in% installed.packages()) install.packages("ggplot2")
library("ggplot2")
if (!"gridExtra" %in% installed.packages()) install.packages("gridExtra")
library("gridExtra")
if (!"caret" %in% installed.packages()) install.packages("caret")
library("caret")

```

##Data Structure and Processing
###Structure
```{r structure}
data(mtcars)
str(mtcars)
```
The dataset shows 32 observations of 11 variables.  The transmission type (am), a focus of the 
analysis, is denoted as either "0" (automatic) or "1" (manual).  In the processing section,
the numeric values of am are translated into factors "automatic" and "manual".

###Processing
```{r processing}
# replace 0&1 with "automatic" and "manual" in new dataframe
mtcars2 <- mtcars
mtcars2$am[mtcars2$am == 0] <- "automatic"
mtcars2$am[mtcars2$am == 1] <- "manual"
```

###Data Exploration
```{r exploration}
meanMPGAuto <- mean(subset(mtcars2, am== "automatic")$mpg)   #mean mpg of automatics
meanMPGManual <- mean(subset(mtcars2, am == "manual")$mpg)   #mean mpg of manuals


```
As apparent in the box plot "Miles per Gallon by Transmission Type" (Appendix - Figure 1), there is a significant difference between the fuel efficiency in miles per gallon (mpg) for automatic and manual transmissions.  Cars with manual transmissions have a mean mpg of `r round(meanMPGManual, 3)`, 
a full `r round(meanMPGManual - meanMPGAuto, 3)` improvement over automatics
(`r round(meanMPGAuto, 3)` mpg). 

##Models
In order to examine this relationship more closely, this analysis will apply a simple linear
regression.  Then, it will look at a multivariable regression model.  The models will be 
compared via nested model testing to determine the best fit.  That fit will be evaluated
using the following methods:

* Residuals Plot
* Q-Q
* Scale-Location
* Leverage

####Simple Linear Regression
```{r slr}
fitS <- lm(mpg ~ am, mtcars2)
summary(fitS)

```
####Multivariable Regression Model
```{r mvrm}
fitM <- lm(mpg ~ am + wt + qsec, mtcars2)
summary(fitM)
```
####Nested Model Testing
```{r nestedmodel}
fit2 <- update(fitS, mpg ~ am + wt)
fit3 <- update(fitS, mpg ~ .)
anova(fitS, fit2, fitM, fit3)
```
###Residual Plot and Diagnostics



###Quantify Uncertainty



##Conclusions


#Appendix
Figure 1:
```{r fig1}
ggplot(aes(x = am, y = mpg), data = mtcars2) +
  geom_boxplot(fill = "blue") +
  labs(x = "Transmission Type", y = "Miles per Gallon") + 
  ggtitle("Miles per Gallon by Transmission Type") + 
  annotate(geom = "text", x = 1, y = meanMPGAuto, 
           color="white", label = "-----", size=7, fontface="bold", 
           angle = 0) + 
  annotate(geom = "text", x = 2, y = meanMPGManual, 
           color="white", label = "-----", size=7, fontface="bold", 
           angle = 0)
```

Figure 2:
```{r fig2}
featurePlot(x = mtcars2[,c(1, 3, 4, 6, 7)], y = mtcars2$am, plot = "pairs", col = c("red", "blue"), 
            title = "Feature Plot for Variables in Analysis")
```

